---
title: "Lab 9 - HPC"
output: 
html_document: default
link-citations: yes
---

# Learning goals

In this lab, you are expected to practice the following skills:

- Evaluate whether a problem can be parallelized or not.
- Practice with the parallel package.
- Use Rscript to submit jobs.

```{r eval=FALSE, echo=FALSE}
# install any missing packages
install.packages("microbenchmark")
```

## Problem 1

Give yourself a few minutes to think about what you learned about parallelization. List three
examples of problems that you believe may be solved using parallel computing,
and check for packages on the HPC CRAN task view that may be related to it.

1. Multidimensional computations (i.e. vector algebra, matrix multiplication)
2. Machine learning training (i.e. hyperparameter tuning, cross-validation)
3. Big data analytics

## Problem 2: Pre-parallelization

The following functions can be written to be more efficient without using
`parallel`:

1. This function generates a `n x k` dataset with all its entries having a Poisson distribution with mean `lambda`.

```{r p2-fun1, eval = FALSE}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n * k, lambda), nrow = n, ncol = k)
}

# Benchmarking
microbenchmark::microbenchmark(
  fun1(),
  fun1alt()
)
```

How much faster?

Since we are generating all Poisson random numbers at once instead of looping n x k times, then it is going to be faster by the order of n x k. In this case its faster by 613 ms on average.


2.  Find the column max (hint: Checkout the function `max.col()`).

```{r p2-fun2, eval = FALSE}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  x_transposed <- t(x)
  max_indices <- max.col(x_transposed)
  x_transposed[cbind(seq_along(max_indices), max_indices)]
}

# Benchmarking
set.seed(1234)
x <- matrix(rnorm(1e4), nrow = 10)


microbenchmark::microbenchmark(
  fun2(x),
  fun2alt(x),
  times = 1000 
)
```

(The plot is given above) It is faster by 991 ms on average


## Problem 3: Parallelize everything

We will now turn our attention to non-parametric 
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)).
Among its many uses, non-parametric bootstrapping allow us to obtain confidence
intervals for parameter estimates without relying on parametric assumptions.

The main assumption is that we can approximate many experiments by resampling
observations from our original dataset, which reflects the population. 

This function implements the non-parametric bootstrap:

```{r p3-boot-fun, eval = FALSE}
library(parallel)

my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
 
  # Making the cluster using `ncpus`
  # STEP 1: GOES HERE
  cl <- makeCluster(ncpus)
  # STEP 2: GOES HERE
  clusterExport(cl, varlist = c("dat", "stat", "idx"), envir = environment())
  
  # STEP 3: THIS FUNCTION NEEDS TO BE REPLACED WITH parLapply
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[, i], , drop = FALSE])
  })
  
  # Coercing the list into a matrix
  ans <- do.call(rbind, ans)
  
  # STEP 4: GOES HERE
  stopCluster(cl)
  
  ans <- do.call(rbind, lapply(ans, function(x) matrix(x, nrow = 1)))
  
  ans
}
```

1. Use the previous pseudocode, and make it work with `parallel`. Here is just an example for you to try:

```{r p3-test-boot, eval = FALSE}
# Bootstrap of a linear regression model
my_stat <- function(data) {
  fit <- lm(y ~ x, data = data)
  coef(fit)
}

# DATA SIM
set.seed(1)
n <- 500 
R <- 1e4
x <- cbind(rnorm(n)) 
y <- x*5 + rnorm(n)
dat <- data.frame(x = x, y = y)

# Check if we get something similar as lm
ans0 <- confint(lm(y ~ x, data = dat))
ans1 <- my_boot(dat, my_stat, R = R, ncpus = 4)


```

2. Check whether your version actually goes faster than the non-parallel version:

```{r benchmark-problem3, eval = FALSE}
library(microbenchmark)

my_boot_nonparallel <- function(dat, stat, R) {
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow = n, ncol = R)
  ans <- lapply(seq_len(R), function(i) {
    stat(dat[idx[, i], , drop = FALSE])
  })
  ans <- do.call(rbind, ans)
  ans
}

# Benchmarking
set.seed(1)
microbenchmark::microbenchmark(
  parallel = my_boot(dat, my_stat, R = R, ncpus = 4),
  non_parallel = my_boot_nonparallel(dat, my_stat, R = R),
  times = 10
)

```

The parallelized version does run faster than the non-parallelized one, around 2 times faster on average.

## Problem 4: Compile this markdown document using Rscript

Once you have saved this Rmd file, try running the following command
in your terminal:

```bash
Rscript --vanilla -e 'rmarkdown::render("[lab09-hpc.Rmd]")' &
```

Where `[full-path-to-your-Rmd-file.Rmd]` should be replace with the full path to
your Rmd file... :).


